---
title: "Applied Analytics Assignment"
author: "Timothy Ndung'u"
date: "2023-10-20"
output: pdf_document
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Reading in the Data

```{r}
library(readr)
stroke_data <- read_csv("C:/Users/Ndungu/Downloads/stroke prediction (edited).csv")
head(stroke_data)

```
# Exploratory Data Analysis (EDA)

## Overview of Data
```{r}
class(stroke_data)
```
```{r}
summary(stroke_data)
```
```{r}
table(stroke_data$stroke)

# 1 = stroke
# 0 = no stroke
```


## Sample Data Plots

```{r}

par(mfrow = c(2,2))

# histogram of target variable (stroke)
hist(stroke_data$stroke, main = "Histogram of Individuals with stroke in stroke prediction", xlab = "stroke")

# frequency plots of character variables 
plot(as.factor(stroke_data$gender), main = "Plot of gender of individuals in stroke prediction", xlab = "gender")
plot(as.factor(stroke_data$ever_married), main = "Plot of individuals ever married in stroke prediction", xlab = "ever_married")
plot(as.factor(stroke_data$work_type), main = "Plot of type of work of individuals in stroke prediction", xlab = "work_type")
plot(as.factor(stroke_data$Residence_type), main = "Plot of type of residence of individuals in stroke prediction", xlab = "residence_type")
plot(as.factor(stroke_data$smoking_status), main = "Plot of smoking status of individuals in stroke prediction", xlab = "smoking_status")

# histogram of numeric variables 
hist(stroke_data$id, main = "Histogram of id no. of individuals in stroke prediction", xlab = "id")
hist(stroke_data$age, main = "Histogram of age of individuals in stroke prediction", xlab = "age")
hist(stroke_data$hypertension, main = "Histogram of individuals with hypertension in stroke prediction", xlab = "hypertension")
hist(stroke_data$heart_disease, main = "Histogram of individuals with heart disease in stroke prediction", xlab = "heart_disease")
hist(stroke_data$avg_glucose_level, main = "Histogram of avg. glucose level of individuals in stroke prediction", xlab = "avg_glucose_level")
hist(stroke_data$bmi, main = "Histogram of bmi of individuals in stroke prediction", xlab = "bmi")

```




# Data Preprocessing

## Character to Numeric Data

Regression models necessitate that the data should be numeric. Otherwise, we need to implement a classification models such as K-Means, KNN, Tree-based methods. 

```{r}

# Factor out numeric columns
num_data <- stroke_data[ ,1:8]
head(num_data )


```

```{r}

# function to change categorical character data to numeric data

gender <- factor(stroke_data$gender, ordered = TRUE)
head(gender)

head(as.numeric(gender))

```

```{r}

# generalizing the function to apply to all columns

char_cols <- stroke_data[ , 9:ncol(stroke_data)]

fact_function <- function(x){
  return(as.numeric(factor(x, ordered = TRUE)))
}

char_data <- lapply(char_cols, fact_function) # returns a list

# sapply() - returns a vector or a matrix

char_data <- as.data.frame(char_data, col.names = colnames(char_cols))

head(char_data)
```

```{r}
# the final dataset

data <- cbind.data.frame(num_data, char_data)
head(data)

```



# Fitting a Logistic Regression Model 

This is given that our response variable is binary where 1 represents default event, while 2 (now changed to 0) represents no default. 

## Train - Test Split


```{r}
# install.packages("caTools") # for logistic regression
# install.packages("ROCR") # for model evaluation

library(caTools)
library(ROCR)
```
```{r}

logi_data <- data

logi_data$stroke <- ifelse(logi_data$stroke == 2, 0, logi_data$stroke)

## assigning default events (2) to value 0 for fitting a logistic model

head(logi_data)

```


```{r}
split_data <- sample.split(logi_data, SplitRatio = 0.7)

# split ration is 70:30

head(split_data)
```
```{r}
traindata <- subset(logi_data, split_data == "TRUE")  # picks datapoints that return TRUE
testdata <- subset(logi_data, split_data == "FALSE")

```


### Fitting the Logistic Model

```{r}

# Logistic Model on the Training model
logistic_model <- glm(traindata$stroke~., data = traindata, family = "binomial")
summary(logistic_model)

```

The dispersion parameter is a scaling parameter added to help fit the model. 

Fisher scoring is a statistical method for finding the maximum likelihood estimate of a parameter.

```{r}

train_predict <- predict(logistic_model, testdata, type = "response")

# predict future probability values using the testdata

head(train_predict)
```

```{r}

# change output to probabilities by assigning them to values 0 or 1
predicted_probs <- ifelse(train_predict >0.5, 1, 0)

head(predicted_probs)
```
```{r}
table(predicted_probs)
table(testdata$stroke)
```


```{r}

# output comparison of predicted probabilities against actual data

output <- data.frame(testdata$stroke, predicted_probs)

head(output)
```

```{r}
library(caret)
predicted = predict(logistic_model,testdata,type="response")
testdata$traindata$stroke = ifelse(testdata$traindata$stroke=="Yes",1,0)
testdata$traindata$stroke = as.factor(testdata$traindata$stroke)
predicted = as.factor(predicted)


```





